---
title: Tidying what should have never been untied. Collective Action, Capitalism, and Social Movements.

# set to 1 to the authors and acknowledgements, producing the required blinded version.
blinded: 0

authors: 
- name: Álvaro Cabrera
  thanks: The author gratefully acknowledge Gay Seidman and Pablo Pérez-Ahumada for their help landing this piece from the world of ideas to the material world.
  affiliation: LACIS, University of Wisconsin-Madison


keywords:
- collective action
- social movements
- capitalism
- class structure
- class politics
abstract: |
 Political processes, social movements, and collective action inquiry have a large and long tradition in social science research. However, most of these studies have relied on traditional approaches to address such research, ignoring the effects that capitalism has on politics, political action, social movements, and collective action. This phenomenon has been repeated all over the globe, particularly in the Global North, in which repertories of mobilization, political processes, and framing analyses have dominated the scholarly debate around collective action. Although some theories have been influenced by Marxism, none of them take capitalism and, thus, class as central in their analyses. Drawing on Marxist collective action theory, Erik Wright’s class model analysis, multivariate regression analysis, and the ISSP Survey (ISSP 2014 – Citizenship II, N = 49.807), the purpose of this paper is to examine the influence that class structure has on participation in different activities of collective action. In consonance with the theory, the results show that belonging to the working class has a negative effect on participation in collective action. Conversely, the upper classes show positive effects in collective action. Lastly, belonging to middle groups doesn’t show any significant effect.

bibliography: references.bib
biblio-style: apalike # default is using plain.bst
output: rticles::asa_article
---

# Introduction

a

# Social Movements, Capitalism, and Class

## Social Movement classic research agenda

Social movement research has a long-standing tradition in academia. Its development, some argue, started from what scholars have called the "psychological" approach [@eidlinSocialClassSocial2018]. Drawing on psychological theory, this approach focused on the individual and sought to understand the reasons that led actors to take part in mobilization. To put it in other words, this analytical framework delved into the internal factors that caused actors to participate and, thus, create social movements.

Drawing on the limits of this approach, the resource mobilization theory quickly gained adepts within academia, particularly in America, to explain that social movements didn't occur simply due to individual deviations but depended on the resources that participants held [@mccarthyResourceMobilizationSocial1977]. As such, actors that were able to mobilize a varied array of resources had the capacity to maintain continuous participation in movements.

This perspective, although criticized, remained dominant for a long period of time, even serving as a background for contemporary academic work. Nonetheless, during the 1980s and influenced by Marxist tradition, a new theoretical framework developed and served as a milestone for social movement scholars. The political process model [@mcadamPoliticalProcessDevelopment1999], McAdam argues, is an alternative to resource mobilization theory and the classical model (psychological). In an attempt to synthesize the social movement theory to date, the theory proposes that insurgency occurs when internal and external factors develop collective action.

In McAdam's words @mcadamPoliticalProcessDevelopment1999,

[@mcadamDynamicsContention2004]

## (Re)turning to Capitalism

The classical social movement agenda led the research on diverse aspects and perspectives on social movement inquiry. However, as Marxist scholars have always argued, capitalism as a social structure is always creating antagonistic forces that clash against each other, and this has been demonstrated by the various conflicts throughout the world that claim a reorder of the economic and political balance. The #OccupyWallstreet movement, the anti-austerity movements in Europe [@dellaportaSocialMovementsTimes2020], and the anti-neoliberal movements in South American countries [@sommaNoWaterOasis2020] have prompted many scholars to return to including capitalism in social movements analysis.

Regarding the European scenario, @dellaportaSocialMovementsTimes2020 claim that recent European state policies have prompted heterogeneous organizations, movements, and other groups to walk to the street against such policies. In such a scenario, the authors argue that

, although academia has left the political economy since 1970, it su

## Class and Social Movements

## Plot

1

```{r, fig.cap = "A plot example", echo = FALSE}
plot(cars)
```

# Methodology

## Data

This article uses a quantitative methodology based on the [Social Survey Programme 2014 - Citizenship II (ISSP)](https://www.gesis.org/en/issp/modules/issp-modules-by-topic/citizenship/2014)[^1] (*N* = 49.087). This dataset applies a similar[^2] questionnaire in every country in which it is applied. The sampling procedure differs for each country: in some countries, partly simple random samples were applied, whereas in others, partly multistage stratified random samples. The data was collected in two ways. The first one was through interviews (face-to-face, Computer Assisted Personal interviews, or completed on the telephone) or Self-administered questionnaires (Computer Assisted self-interviews or Computer-Assisted web interviews). The analysis is mostly applied to individuals that are 18 years or more with some exceptions[^3]. After processing the dataset, the sample was reduced to 33.582 individuals nested in 34 countries.

[^1]: By the time this article has been written, another wave of the same thematic survey is under development and expected to fully realese in 2025.

[^2]: It is a similar questionary because the survey don't apply the exact same questions in every country. However, the questions are prepared to grasp and collect the same dimensions for every question applied.

[^3]: According to the ISSP, the exceptions were countries such as Finland, were individuals between 15 and 75 were surveyed, Japan, were 16 years old and older were surveyed, South Africa, were the respondents were 16 years or more, and Sweden, were the individuals were between 17 and 79 years old.

## Methods

[@huntington-kleinEffectIntroductionResearch2022]

The hypotheses that led this study were tested by calculating OLS. According to Linear regressions are the "*best linear approximation of the relationship*" between two variables.

As the literature says, Linear Models are appropriate when we seek [....]. In this case, since we are testing the effect that class structure has on the working class participation in collective action activities, other models that provide a better understanding of probability, such as Logistic and Logit, aren't required. To put it in other words, since we are not measuring the *probability* that the working class has to undertake collective action activities but the *influence* that class structure itself has on

# Analysis

Prior to the statistical analysis, all the coding workflow was done using R and its many packages[^4].

[^4]: Most of the coding, wrangling, and analyses were used with tidyverse[@tidyverse], haven [@haven], sjPlot [@sjPlot], sjmisc[@sjmisc], sjlabelled [@sjlabelled], summarytools [@summarytools], car [@car], openxlsx[@openxlsx]. For presentation purposes stargazer [@stargazer] and xtable [@xtable] were used.

## Descriptives

```{=tex}
\begin{table}[ht]
\centering
\caption{Descriptives Statistics}
\begin{tabular}{rllrrrrl}
  \hline
 & var & label & n & NA.prc & mean & sd & range \\ 
  \hline
3 & acc & Collective Action Participation & 33582 & 8.51 & 1.39 & 1.32 & 4 (0-4) \\ 
  4 & acc2 & Dummy Collective Action & 33582 & 8.51 & 0.65 & 0.48 & 1 (0-1) \\ 
  6 & css & Social Class & 36706 & 0.00 & 6.67 & 2.45 & 8 (1-9) \\ 
  5 & cs\_sub & Subjective Social-Class & 34224 & 6.76 & 0.96 & 0.74 & 2 (0-2) \\ 
  1 & AGE & Age of respondent & 36627 & 0.22 & 48.86 & 16.67 & 87 (15-102) \\ 
  2 & SEX & Sex of Respondent & 36696 & 0.03 & 1.51 & 0.50 & 1 (1-2) \\ 
  7 & partner & Living in steady partnership & 35142 & 4.26 & 0.69 & 0.46 & 1 (0-1) \\ 
   \hline
\end{tabular}
\end{table}
```
## Models

```{r, fig.cap = "A plot example", echo = FALSE}

```

# Conclusion {#conc}

# References
