---
title: "Tidying what never should have been untied: Social Movements, Capitalism, and Class"
subtitle: "LACIS, University of Wisconsin-Madison."
author: "Álvaro Cabrera"
date: "`r Sys.Date()`"
abstract: "Political processes, social movements, and collective action inquiry have a large and long tradition in social science research. However, most of these studies have relied on traditional approaches to address such research, ignoring the effects that capitalism has on politics, political action, social movements, and collective action. This phenomenon has been repeated all over the globe, particularly in the Global North, in which repertories of mobilization, political processes, and framing analyses have dominated the scholarly debate around collective action. Although some theories have been influenced by Marxism, none of them take capitalism and, thus, class as central in their analyses. Drawing on Marxist collective action theory, Erik Wright’s class model analysis, multivariate regression analysis, and the ISSP Survey (ISSP 2014 – Citizenship II, N = 49.807), the purpose of this paper is to examine the influence that class structure has on participation in different activities of collective action. In consonance with the theory, the results show that belonging to the working class has a negative effect on participation in collective action. Conversely, the middle groups show positive effects in collective action. Lastly, belonging to upper class shows a negative effect."

output:
  pdf_document:
    latex_engine: pdflatex
    includes:
      in_header: preamble.tex
bibliography: references.bib
csl: apa7.csl
link-citations: yes
---

\newpage

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE)
library(tinytex)
```

```{=pdf}
<style>
body {
  text-align: justify;
}
</style>
```
# Introduction

Social movements scholars have been researching collective action for a long period of time. Based mainly upon traditional perspectives, such as resource mobilization, political processes, political opportunities, framing, and contentious politics, the output of social has been proficient until these days. Answering questions such as "who" joins social movements, who leads them, what are the emotions associated when protesting, why are some violent rather than pacific, or why do social movements don't achieve anything, whereas others influence public policy, has been at the core of this analytic framework.

Although undoubtedly prolific, the rising of social movements that mainly focused on what Marxist tradition calls "the structure," that is, the socio-economic relations. From the #Occupy Wall Street movement demanding more rights for the 99% in the U.S., the *Indignados* in Spain occupying public spaces and demanding new social pacts, the recent anti-austerity movements in Europe demanding the end of the public cuts, to the recent anti-neoliberal revolts in South America,

By the time these uprisings have been unfolding, scholars have shown a growing interest in the foundations that could explain these demonstrations. This has led to a revitalization of political economy, class, and labor in explaining the origins of movements. This article is an attempt to contribute to such a purpose.

Although sometimes influenced by Marxist tradition [@mcadamPoliticalProcessDevelopment1999; @mcadamDynamicsContention2004; @tarrowPowerMovementSocial2011], the classic social movement agenda[^1] has somehow omitted the central role of capitalism in the making of social movement. In this work, however, I focus on the effects of class structure as a central component in explaining social movements. Drawing on contemporary Marxist theory and logistic multivariate regression models, I argue that class structure negatively impacts working-class participation in collective action activities.

[^1]: @mcadamDynamicsContention2004 refer to political opportunities theory, mobilizing resource, and collective action frames as the classical social movement agenda. I argue that contentious politics is, as well, in such a list.

In what follows, the rest of the paper is organized into five major parts. The first part corresponds to the literature review, where major bibliographic and prior research on social movements, political economy and social movements, and class and social movements is reviewed. The second part corresponds to the methodology. In this part, the methods used and the data are described. This section is followed by the analysis section, where the hypothesis[^2] is presented, and the statistical model is calculated and plotted. The discussion section follows the analysis one, where the results are discussed in light of literature and prior research. Finally, the last section corresponds to the conclusion.

[^2]: It is very likely that, as this research goes forward, this unfolds into more than one hypothesis.

# Social Movements, capitalism, and class

## Social Movement classic research agenda

Social movement studies have a long-standing tradition of studying conflict, social change, and what some scholars contentious politics.

However, with the recent protests located in the Global North have shown that

Social movement research has a long-standing tradition in academia. Its development, some argue, started from what scholars have called the "psychological" approach [@eidlinSocialClassSocial2018]. Drawing on psychological theory, this approach focused on the individual and sought to understand the reasons that led actors to take part in mobilization. To put it in other words, this analytical framework delved into the internal factors that caused actors to participate and, thus, create social movements.

Drawing on the limits of this approach, the resource mobilization theory quickly gained adepts within academia, particularly in America, to explain that social movements didn't occur simply due to individual deviations but depended on the resources that participants held [@mccarthyResourceMobilizationSocial1977]. As such, actors that were able to mobilize a varied array of resources had the capacity to maintain continuous participation in movements.

This perspective, although criticized, remained dominant for a long period of time, even serving as a background for contemporary academic work. Nonetheless, during the 1980s and influenced by Marxist tradition[^3], a new theoretical framework developed and served as a milestone for social movement scholars. The political process model [@mcadamPoliticalProcessDevelopment1999], McAdam argues, is an alternative to resource mobilization theory and the classical model (psychological). In an attempt to synthesize the social movement theory to date, the theory proposes that insurgency occurs when internal and external factors develop collective action.

[^3]: McAdam explicitly uses Marxist framework to develop his theory. Also, @tarrowPowerMovementSocial2011 recognizes the importance of the Marxist tradition of collective Action. For more details see @hetlandStrangeDisappearanceCapitalism2013.

In McAdam's words @mcadamPoliticalProcessDevelopment1999,

[@mcadamDynamicsContention2004]

## (Re)turning to Capitalism

The classical social movement agenda led the research on diverse aspects and perspectives on social movement inquiry. However, as Marxist scholars have always argued, capitalism as a social structure is always creating antagonistic forces that clash against each other, and this has been demonstrated by the various conflicts throughout the world that claim a reorder of the economic and political balance. The #OccupyWallstreet movement, the anti-austerity movements in Europe [@dellaportaSocialMovementsTimes2020], and the anti-neoliberal movements in South American countries [@sommaNoWaterOasis2020] have prompted many scholars to return to including capitalism in social movements analysis.

Regarding the European scenario, @dellaportaSocialMovementsTimes2020 claim that recent European state policies have prompted heterogeneous organizations, movements, and other groups to walk to the street against such policies. In such a scenario, the authors argue that

, although academia has left the political economy since 1970, it su

## Class and Social Movements

# Methodology

## Data

This article uses a quantitative methodology based on the [Social Survey Programme 2014 - Citizenship II (ISSP)](https://www.gesis.org/en/issp/modules/issp-modules-by-topic/citizenship/2014)[^4] (*N* = 49.087). This dataset applies a similar[^5] questionnaire in every country in which it is applied. The sampling procedure differs for each country: in some countries, partly simple random samples were applied, whereas in others, partly multistage stratified random samples. The data was collected in two ways. The first one was through interviews (face-to-face, Computer Assisted Personal interviews, or completed on the telephone) or Self-administered questionnaires (Computer Assisted self-interviews or Computer-Assisted web interviews). The analysis is mostly applied to individuals that are 18 years or more with some exceptions[^6]. After processing the dataset, the sample was reduced to 33.582 individuals nested in 34 countries.

[^4]: By the time this article has been written, another wave of the same thematic survey is under development and expected to fully realese in 2025.

[^5]: It is a similar questionary because the survey don't apply the exact same questions in every country. However, the questions are prepared to grasp and collect the same dimensions for every question applied.

[^6]: According to the ISSP, the exceptions were countries such as Finland, were individuals between 15 and 75 were surveyed, Japan, were 16 years old and older were surveyed, South Africa, were the respondents were 16 years or more, and Sweden, were the individuals were between 17 and 79 years old.

## Methods

[@huntington-kleinEffectIntroductionResearch2022]

The hypotheses that led this study were tested by calculating OLS. According to Linear regressions are the "*best linear approximation of the relationship*" between two variables.

As the literature says, Linear Models are appropriate when we seek [....]. In this case, since we are testing the effect that class structure has on the working class participation in collective action activities, other models that provide a better understanding of probability, such as Logistic and Logit, aren't required. To put it in other words, since we are not measuring the *probability* that the working class has to undertake collective action activities but the *influence* that class structure itself has on

# Analysis

Prior to the statistical analysis, all the coding workflow was done using R and its many packages[^7].

[^7]: Most of the coding, wrangling, and analyses were used with tidyverse [@tidyverse], haven [@haven], sjPlot [@sjPlot], sjmisc [@sjmisc], sjlabelled [@sjlabelled], summarytools [@summarytools], car [@car], openxlsx [@openxlsx]. For presentation purposes stargazer [@stargazer], effects [@effects], and xtable [@xtable] were used.


The results of the codification can be seen in Table \ref{desc-table}

```{=tex}
\begin{table}[ht]
\centering
\caption{Descriptives Statistics}
\label{desc-table}
\begin{tabular}{rllrrrrl}
  \hline
 & var & label & n & NA.prc & mean & sd & range \\ 
  \hline
3 & acc & Collective Action Participation & 33582 & 8.51 & 1.39 & 1.32 & 4 (0-4) \\ 
  4 & acc2 & Dummy Collective Action & 33582 & 8.51 & 0.65 & 0.48 & 1 (0-1) \\ 
  6 & css & Social Class & 36706 & 0.00 & 6.67 & 2.45 & 8 (1-9) \\ 
  5 & cs\_sub & Subjective Social Class & 34224 & 6.76 & 0.96 & 0.74 & 2 (0-2) \\ 
  1 & AGE & Age of respondent & 36627 & 0.22 & 48.86 & 16.67 & 87 (15-102) \\ 
  2 & SEX & Sex of Respondent & 36696 & 0.03 & 1.51 & 0.50 & 1 (1-2) \\ 
  7 & partner & Living in steady partnership & 35142 & 4.26 & 0.69 & 0.46 & 1 (0-1) \\ 
   \hline
\end{tabular}
\end{table}
```
## Hypotheses

### Hypothesis 1

According to the Marxist tradition, social struggle, or insurgency, comes from the antagonistic relations that class structure creates. Therefore, individuals who are not in possession of the means of production, the working class, tend to resist and sometimes rebel against those who own them. [@marxManifiestoPartidoComunista2019; @marxDieciochoBrumarioLuis2015; @marxAntologia2015; @marxContribucionCriticaFilosofia2019]. Nonetheless, many critics have risen due to the apparent "mechanical" way in which such a process would unfold. In Marx's traditional view, capitalist development would generate the conditions in which the proletariat, the revolutionary class, would rebel against the bourgeoisie in a definitive attempt to suppress class structure and unequal relations. This theoretical framework inspired, for the first time in history, social struggles all over the world in which the main actor was the working class [@hobsbawmHistoriaSigloXX2011]. Nonetheless, although some of these revolutions[^8] still remain, the great majority of them collapsed and faded away.

[^8]: Chinese, Cuban, and Vietnamese revolutions are yet developing in a rather traditional view of such a tradition.

In an attempt to prove that Marxist theory is yet invaluable in examining capitalism, some scholars have clarified the position of antagonistic relations and collective action. Drawing on the contribution of @gramsciSelectionsPrisonNotebooks1971 and @poulantzasClaseSocialesCapitalismo1990, a new approach to Marxist collective action has been developing [@wrightInterrogatingInequalityEssays1994; @wrightClasses1985; @wrightHowBeAnticapitalist2019; @chibberClassMatrixSocial2022; @chibberConfrontingCapitalism2022; @chibberPostcolonialTheorySpecter2014; @hetlandStrangeDisappearanceCapitalism2013; @eidlinSocialClassSocial2018; @przeworskiCapitalismSocialDemocracy1985].

The role of collective action, insurgency, and social conflicts stills occupy a central role in this tradition. The difference, however, is in what @chibberClassMatrixSocial2022 calls capitalism's class structure durability and survival capacity. If, in the classic Marxist tradition, the development of the economic structure would prompt the working class to become a *class in itself* for a *class for itself* and thusoverthrow the ruling class, in the current Marxist tradition, that classic statement has a new formulation: class structure does not prompt antagonistic collective action, but individualistic responses to economic malaise.

Following Gramsci's groundbreaking ideas of consent, @przeworskiCapitalismSocialDemocracy1985, in his classic text, was one of the first to try to openly operationalize why class struggle wasn't developing all over the world. Recognizing the limited options that class structures provide to those who oppose it, in his analysis, he came to the conclusion that capitalist reproduction has a material basis of consent which is, at the same time, dependant on wages. 

**Hypothesis 1 (H1)**: In contemporary capitalism, the class structure itself influences the working class to be resigned. Thus, the working class is less likely to partake in collective action activities.

## Models

As shown in Table \ref{modelo-a}


```{r loading-model-a, include=FALSE}

library(stargazer)

load("C:/Users/Alvaro C/Dropbox/3. Educacion/1. MA_LACIS/Lacis_MA/4. Data_Tables/modelo_a_star.RData")

```

```{r printing-model-a, echo=FALSE, results='asis'}

cat(modelo_a_star)

```

# Discussion

```{r plotting-model-a, include=FALSE}
library(effects)

db2<-readRDS("C:/Users/Alvaro C/Dropbox/3. Educacion/1. MA_LACIS/Lacis_MA/3. Dataframes/2_analyses.RDS")
load("C:/Users/Alvaro C/Dropbox/3. Educacion/1. MA_LACIS/Lacis_MA/5. Plots/reg1.RData")

```

```{r plotting-model-a-2, echo=FALSE, fig.cap="Logistic regression predicted values"}

plot(allEffects(reg1)) #Plot modelo A


```

# Conclusions

# References
